Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-01 Manuscript/manuscriptPS.tex
Tuned the abstract.

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
electrostatic interactions and direct binding, and can induce
membrane fusion and phase separation in the presence of calcium ions.
Molecular details of these phenomena are not well understood,
because accurate models to interpret the experimental data have not
been available. Here, we use solid-state NMR S-DROSS to measure the signs of the experimental POPS headgroup
order parameters, and gather a set of experimental NMR results from pure PS and mixed PC:PS lipid bilayers. Using the open collaboration
method, we then extract data from a wide range of available PS MD models (force fields) and asses the PS headgroup structures and the ion binding produced by these force fields by comparing to the NMR data. We find that none of the models reproduce the NMR data within experimental accuracy,
but the best ones suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity---the response of PS headgroups to bound ions can even
\emph{qualitatively} differ from experiments. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

>>>>

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus S. Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
nonspecific electrostatic interactions as well as specific binding; in the presence of calcium ions, PS can induce
membrane fusion and phase separation.
Molecular details of these phenomena remain not well understood, partly
because accurate models to interpret the experimental data have not
been available. Here we
gather a set of experimental NMR data of C--H bond order parameters, $S_\mathrm{CH}$, for pure
PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
and augment this data set by measuring with solid-state NMR S-DROSS the signs of $S_\mathrm{CH}$ in the POPS headgroup.
We then use this data set
to assess,
in a wide range of available molecular dynamics (MD) simulation models (force fields) of PS,
the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes.
We find that none of the MD models reproduce the NMR data within experimental accuracy.
The best MD models, however, suggest that the carboxyl group in the PS headgroup
does not rotate freely. In line with our previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity; moreover, the response of PS headgroups to bound ions can  differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}
